---
title: "codling data analysis"
output: html_document
date: "2024-10-25"
---

## Introduction

The research that generated these data was in part funded by New Zealand pipfruit growers. The published analysis was funded by New Zealand pipfruit growers.
Data are from trials that studied the mortality response of codling moth to fumigation with methyl bromide.

```{r}
# libraries
library(tidyverse)
library(brms)
library(tidybayes)
library(marginaleffects)
library(ggthemes)
library(ggsci)
library(cowplot)
library(patchwork)

# Import dataset
data <- read_csv(paste(getwd(), "/codling.csv", sep = ""))
```


### Variables explanation

  + `dose`: injected dose of methyl bromide, in mg per cubic meter 
  + `tot`: number of insects in chamber
  + `dead`: number of insects dying
  + `pobs`: proportion dying
  + `cm`: control mortality (i.e., at dose 0)
  + `ct`: concentration-tim sum
  + `Cultivar`: a factor with levels `BRAEBURN`, `FUJI`, `GRANNY`, `Gala`, `ROYAL`, `Red Delicious`, and `Splendour`
  + `gp`: a factor which has a different level for each different combination of `Cultivar`, `year`, and `rep` (replicate)
  + `year`: a factor with levels 1988 and 1989
  + `numcm`: total number of control insects
  
  
## Exploratory Data Analysis (EDA) 

```{r}
data %>%
  ggplot(aes(dose, pobs)) +
  geom_point(alpha = 0.7, aes(col = factor(year), size = numcm)) +
  facet_wrap(~ Cultivar) +
  geom_smooth(se = FALSE, col = "black") +
  labs(x = "Dose (mg)",
       y = "Proportion of insects dying",
       col = "Year",
       size = "Total number of control insects") +
  theme_clean() +
  theme(legend.position = "bottom") +
  scale_color_aaas() +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  scale_y_continuous(breaks = seq(0, 1, 0.25))
```


The majority of dose-response patterns follow a type of sigmoid, which can be achieved using the **beta** likelihood.


## Fitting our models 

```{r}
# linear model
model_1 <- brm(pobs ~ 1 + Cultivar + dose,
               data = data,
               family = gaussian, # Gaussian likelihood
               chains = 4,
               iter = 2000,
               seed = 25102024)

# beta model
model_2 <- brm(pobs ~ 1 + Cultivar + dose,
               data = data,
               family = Beta, # Beta likelihood
               chains = 4,
               iter = 2000,
               seed = 25102024)
```



### Posterior predictive checking

```{r}
set.seed(2510)

pp_check(model_1, type = "dens_overlay_grouped", group = "Cultivar", ndraws = 100)


pp_check(model_2, type = "dens_overlay_grouped", group = "Cultivar", ndraws = 100)

```



## Predictions

```{r}
# data grid creation for predictions
newdata <- crossing(Cultivar = unique(data$Cultivar),
                    dose = seq(0, 30))


# which likelihood will produce more confident predictions?
bind_rows(
          # compute the gaussian fitted line
          fitted(model_1, newdata = newdata) %>%
            data.frame() %>%
            bind_cols(newdata),
          
          # compute the beta fitted line
          fitted(model_2, newdata = newdata) %>%
            data.frame() %>%
            bind_cols(newdata)) %>%
  
  # data wrangling
  mutate(likelihood = rep(c("Gaussian", "Beta"), each = n()/2) %>% factor(levels = c("Gaussian", "Beta"))) %>%
  
  # plot
  ggplot(aes(x = dose)) +
  geom_hline(yintercept = 0:1, color = "white", linetype = 2) +
  geom_vline(xintercept = c(0, 30), color = "white", linetype = 2) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5, fill = likelihood),
              linewidth = 0, alpha = 0.5) +
  geom_point(data = data,
             aes(y = pobs)) +
  scale_fill_viridis_d(option = "B", begin = 0.15, end = 0.8) +
  facet_wrap(~ Cultivar) +
  labs(x = "Dose (mg)",
       y = "Proportion of insects dying",
       fill = "Likelihood") +
  theme_dark() +
  theme(legend.position = "bottom")

```


### Marginal effects predictions for both likelihoods

```{r}

# using the linear model
model_1 %>% 
  epred_draws(newdata = expand_grid(Cultivar = unique(data$Cultivar),
                    dose = seq(0, 30))) %>%

ggplot(aes(x = dose, y = .epred)) +
  facet_wrap(~ Cultivar) +
  stat_lineribbon() + 
  scale_fill_brewer(palette = "Purples") +
  labs(x = "Dose (mg)", y = "Proportion of insects dying",
       fill = "Credible interval") +
  theme_pander() +
  scale_y_continuous(limits = c(-0.25, 1.25)) +
  theme(legend.position = "bottom")



# using our beta model
model_2 %>% 
  epred_draws(newdata = expand_grid(Cultivar = unique(data$Cultivar),
                    dose = seq(0, 30))) %>%

# plot
ggplot(aes(x = dose, y = .epred)) +
  facet_wrap(~ Cultivar) +
  stat_lineribbon() + 
  scale_fill_brewer(palette = "Purples") +
  labs(x = "Dose (mg)", y = "Proportion of insects dying",
       fill = "Credible interval") +
  theme_pander() +
  scale_y_continuous(limits = 0:1) +
  theme(legend.position = "bottom")

```


## Average dose-response relationship between treatment dosage and proportion of eliminated insects

```{r}
# using our beta model
model_2 %>% 
  epred_draws(newdata = expand_grid(Cultivar = unique(data$Cultivar),
                    dose = seq(0, 30))) %>%

# plot
ggplot(aes(x = dose, y = .epred)) +
  # facet_wrap(~ Cultivar) +
  stat_lineribbon() + 
  annotate("segment", x = 0, xend = 21.8, y = 0.75, yend = 0.75, col = "black", size = 1.2, linetype = 2) +
  annotate("segment", x = 21.8, xend = 21.8, y = 0, yend = 0.75, col = "black", size = 1.2, linetype = 2) +
  geom_point(aes(x = x, y = y), data = data.frame(x = 21.8, y = 0.75), shape = 13, size = 8) +
  annotate("text", x = 15.5, y = 0.8, label = "Minimal effective dose?", size = 7) +
  scale_fill_brewer(palette = "Purples") +
  labs(x = "Dose (mg)", y = "Proportion of insects dying",
       fill = "Credible interval",
       title = "Average dose-response relationship",
       subtitle = "A treatment is considered effective when it eliminates, at least, 75% of the insects") +
  theme_pander() +
  scale_y_continuous(limits = 0:1) +
  theme(legend.position = "bottom")
```


### Which is the optimal treatment?

```{r}
sink(file = "output.txt")

# predictions
model_2 %>% 
  epred_draws(newdata = expand_grid(Cultivar = unique(data$Cultivar),
                    dose = seq(0, 30))) %>% 
  group_by(Cultivar, dose) %>%
  
  # computation of the mean and 95% CI of the predictions
  mean_qi(.epred) %>%
  
  # filter by predicted responses above 75% (i.e., the minimal effective threshold)
  filter(.epred > .75) %>% 
  
  # just keep the first line of each treatment (i.e., the minimal effective dose)
  filter(!duplicated(Cultivar)) %>%
  as.data.frame()

sink()
```

```{r, echo=FALSE}
sessionInfo()
```

